\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-05 paper/ms.tex
Made predicted time notation consistent

For predicting or interpolating missing or future data points (and their variance) at times $\bvec{t_*} = \left\{t_{*,1},t_{*,2},...,t_{*,M}\right\}$, we need to apply the covariance between the measured and predicted times.  The mean of the predicted values is (equation 2.19 from Rasmussen \& Williams):
\begin{eqnarray}
\bar f_* &=& K(\bvec{t_*},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y},\\
&=& K(\bvec{t_*},\bvec{t}) \bvec{z}
\end{eqnarray}
where $\bvec{y}$ is the vector of measurements and $\sigma$ is the measurement uncertainty, and
$\bvec{z} = \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y}$ can be computed from
the Cholesky decomposition of the full covariance matrix in ${\cal O}(J^2N)$.

Since the covariance matrix may be written as the sum of exponentials, the predicted mean is given by
\begin{eqnarray}
\bar f_{*,m} &=& K(t_{*,m},\bvec{t}) \bvec{z} \\
&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t_{*,m}-t_n\vert}\left(a_j \cos{(d_j \vert t_{*,m}-t_n\vert)} +b_j \sin{(d_j \vert t_{*,m}-t_n\vert)}\right) z_j,
\end{eqnarray}
for $m =1$ to $M$.
Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t_{*,m}$ and
$\{t_{n0+1},...,t_N\} \ge t_{*,m}$, gives
\begin{eqnarray}
\bar f_{*,m} &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t_{*,m}-t_n)} \left[a_j \cos{(d_j(t_{*,m}-t_n))}+b_j\sin{(d_j(t_{*,m}-t_n))}\right] z_n \\
&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t_{*,m})} \left[a_j \cos{(d_j(t_n-t_{*,m}))}+b_j\sin{(d_j(t_n-t_{*,m}))}\right] z_n.
\end{eqnarray}
We can compute this recursively.  Defining
\begin{eqnarray}
Q^-_{n_0,i} &=& \left(Q^-_{n_0-1,i}+z_n\tilde V_{n_0,i}\right)e^{-c_j(t_{n+1}-t_n)},\\
X^-_{m,n_0,i} &=& e^{-c_j(t_m^*-t_{n+1})}\tilde U_{m,i},\\
Q^+_{n_0,i} &=& \left(Q^+_{n_0+1,i}+z_n\tilde U_{n,i}\right)e^{-c_j(t_n-t_{n-1})},\\
X^+_{m,n,i} &=& e^{-c_j(t_{n-1}-t_m^*)} \tilde V_{m,i},
\end{eqnarray}
where $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$ to $2J$, and in $X^\pm_{m,n,i}$ the expressions
for $\tilde U_{m,i}$ and $\tilde V_{m,i}$ are evaluted at $t_{m}^*$.
Two sweeps must be carried out, upward in $n$ and $m$ to compute $Q^-$ recursively, and downward in
$m$ and $n$ to compute $Q^+$ recursively, for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.
For each value of $m$, $Q^\pm$ are recursively updated over $n$ until $n_0$ is reached, at which
point the predicted mean can be computed from
\begin{equation}
\bar f_{*,i} = \sum_{i=1}^{2J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0,i}X^+_{m,n_0,i}\right],
\end{equation}

%The variance evaluated at future predicted times $\bvec{t_*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
%\begin{eqnarray}
%\mathrm{var}(f_{*,i+1}) &=& K(t_{*,i+1},t_{*,i+1}) - K(t_{*,i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t_{*,i+1})\\
%&=& K(t_{*,i+1},t_{*,i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} \\
%&\times& e^{-\beta_p(t_{*,i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}\\
%&=& K(t_{*,i+1},t_{*,i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t_{*,i+1}-t_{*,i})} P_{i,p,q}\\
%\end{eqnarray}
%for $1 \le p,q \le J$,
%where
%\begin{equation}
%P_{i,p,q} = e^{-\beta_p(t_{*,i}-\bvec{t}^T)}
%K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t_{*,i}-\bvec{t})}.
%\end{equation}
%For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
%with the Cholesky decomposition.
%Subsequent $P_{i+1,p,q}$ may be obtained recursively as
%\begin{equation}
%P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t_{*,i+1}- t_{*,i})} P_{i,p,q},
%\end{equation}
%in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
%Note that this variance formula {\it only} works for prediction; we have
%yet to derive a formula for which $\bvec{t_*}$ and $\bvec{t}$ are
%interspersed.

>>>>

For predicting or interpolating missing or future data points (and their variance) at times $\bvec{t^*} = \left\{t_{1}^*,t_{2}^*,...,t_{M}^*\right\}$, we need to apply the covariance between the measured and predicted times.  The mean of the predicted values is (equation 2.19 from Rasmussen \& Williams):
\begin{eqnarray}
\bar f_* &=& K(\bvec{t^*},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y},\\
&=& K(\bvec{t^*},\bvec{t}) \bvec{z}
\end{eqnarray}
where $\bvec{y}$ is the vector of measurements and $\sigma$ is the measurement uncertainty, and
$\bvec{z} = \left[K(\bvec{t},\bvec{t})+\sigma^2 I\right]^{-1} \bvec{y}$ can be computed from
the Cholesky decomposition of the full covariance matrix in ${\cal O}(J^2N)$.

Since the covariance matrix may be written as the sum of exponentials, the predicted mean is given by
\begin{eqnarray}
\bar f^*_m &=& K(t^*_m,\bvec{t}) \bvec{z} \\
&=& \sum_n \sum_{j=1}^J e^{-c_j \vert t^*_m-t_n\vert}\left(a_j \cos{(d_j \vert t^*_m-t_n\vert)} +b_j \sin{(d_j \vert t^*_m-t_n\vert)}\right) z_j,
\end{eqnarray}
for $m =1$ to $M$.
Dividing the sum over $n$ into $\{t_1,...,t_{n_0}\} < t^*_m$ and
$\{t_{n0+1},...,t_N\} \ge t^*_m$, gives
\begin{eqnarray}
\bar f^*_m &=& \sum_{j=1}^J\sum_{n=1}^{n_0} e^{-c_j(t^*_m-t_n)} \left[a_j \cos{(d_j(t^*_m-t_n))}+b_j\sin{(d_j(t^*_m-t_n))}\right] z_n \\
&+& \sum_{j=1}^J\sum_{n=n_0+1}^N  e^{-c_j(t_n-t^*_m)} \left[a_j \cos{(d_j(t_n-t^*_m))}+b_j\sin{(d_j(t_n-t^*_m))}\right] z_n.
\end{eqnarray}
We can compute this recursively.  Defining
\begin{eqnarray}
Q^-_{n_0,i} &=& \left(Q^-_{n_0-1,i}+z_n\tilde V_{n_0,i}\right)e^{-c_j(t_{n+1}-t_n)},\\
X^-_{m,n_0,i} &=& e^{-c_j(t_m^*-t_{n+1})}\tilde U_{m,i},\\
Q^+_{n_0,i} &=& \left(Q^+_{n_0+1,i}+z_n\tilde U_{n,i}\right)e^{-c_j(t_n-t_{n-1})},\\
X^+_{m,n,i} &=& e^{-c_j(t_{n-1}-t_m^*)} \tilde V_{m,i},
\end{eqnarray}
where $Q^-_{0,i} = 0$ and $Q^+_{N+1,i}=0$ for $i=1$ to $2J$, and in $X^\pm_{m,n,i}$ the expressions
for $\tilde U_{m,i}$ and $\tilde V_{m,i}$ are evaluted at $t_m^*$.
Two sweeps must be carried out, upward in $n$ and $m$ to compute $Q^-$ recursively, and downward in
$m$ and $n$ to compute $Q^+$ recursively, for a total scaling of $\approx \mathcal{O}(J^2N+J(N+M))$.
For each value of $m$, $Q^\pm$ are recursively updated over $n$ until $n_0$ is reached, at which
point the predicted mean can be computed from
\begin{equation}
\bar f^*_i = \sum_{i=1}^{2J} \left[Q^-_{n_0,i}X^-_{m,n_0,i}+Q^+_{n_0,i}X^+_{m,n_0,i}\right],
\end{equation}

%The variance evaluated at future predicted times $\bvec{t^*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
%\begin{eqnarray}
%\mathrm{var}(f^*_{i+1}) &=& K(t^*_{i+1},t^*_{i+1}) - K(t^*_{i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t^*_{i+1})\\
%&=& K(t^*_{i+1},t^*_{i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t^*_{i+1}-t^*_{i})} \\
%&\times& e^{-\beta_p(t^*_{i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t^*_{i}-\bvec{t})}\\
%&=& K(t^*_{i+1},t^*_{i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t^*_{i+1}-t^*_{i})} P_{i,p,q}\\
%\end{eqnarray}
%for $1 \le p,q \le J$,
%where
%\begin{equation}
%P_{i,p,q} = e^{-\beta_p(t^*_{i}-\bvec{t}^T)}
%K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t^*_{i}-\bvec{t})}.
%\end{equation}
%For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
%with the Cholesky decomposition.
%Subsequent $P_{i+1,p,q}$ may be obtained recursively as
%\begin{equation}
%P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t^*_{i+1}- t^*_{i})} P_{i,p,q},
%\end{equation}
%in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
%Note that this variance formula {\it only} works for prediction; we have
%yet to derive a formula for which $\bvec{t^*}$ and $\bvec{t}$ are
%interspersed.
